Human capital mediates natural selection in contemporary humans

### 2021-12-08 why-natural-selection.Rmd
Tweaks in response to NHB reviews. Rewrite comments on alternative theories.

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals, corrected for ancestry using 100 genetic principal components (see
Materials and Methods). Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores in the sample by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order of 5% of
a standard deviation. These changes could reflect natural selection within the
UK population, but also emigration, or ascertainment bias within the sample.
Respondents have higher income and are better educated than the UK population,
and they may also differ on other unobserved characteristics
[@10.1093/aje/kwx246]. Since richer and more educated people also live longer, this
bias could also increase with age. Lastly, the sample sex ratio skews 
`r female_ratio * 100`% female.

@@

To correct for ascertainment bias, we weight participants using population
data. We try three alternative weighting schemes: (1) "Geographical" weighting,
by geography, age and presence/absence of a partner; (2) age and highest educational
qualification; and (3) for women only, age, highest qualification, and age at 
first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes, and also with uniform
weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
reaching significance under at least one scheme. Weighting makes
a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic variables.

We also check for balancing and diversifying selection by estimating
\@ref(eq:regression) with a quadratic term. In particular, we find diversifying
selection for educational attainment polygenic scores: at higher values of these
scores, the negative effect on fertility is smaller (Appendix Figure
\@ref(fig:plot-purifying)).

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income, education and sex, and
family structure variables including age at first live birth, presence of a partner, and
lifetime number of sexual partners.
We emphasize that these categories are not necessarily exogenous to the polygenic
scores. For example, both in the data and in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related 
polygenic scores. Nevertheless, differences in equation
\@ref(eq:regression) across subgroups constrain the set of possible explanations
for natural selection. As we show below, the economic theory predicts 
the patterns in the data, while alternative explanations do not.

@@

To show this, we write down a simple model of fertility choices. $h$ is
an individual's level of human capital. For now, we simply identify this with
his or her wage $W$. Raising a child takes time $b$. People maximize utility
from the number of children $N$, and income $Y\equiv(1-bN)W$. An individual's
payoff is

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$
in a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical 
predictions, along with corresponding empirical results for the correlation between
polygenic scores and fertility.[^one-period-results]

[^one-period-results]: Predictions 1-3 also hold in the one-period model. Our empirical 
results are actually stronger than prediction 5, in that
correlations with fertility are *reversed* at higher AFLB. This prediction can be
accommodated in the model if children have a money cost as well as a time cost. 
Appendix Figure \@ref(fig:N-plot-with-m) shows an example.

@@

The above does not aim to provide a complete theory of fertility. Rather, it
shows that a relatively simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise.

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels.[^note] Another alternative is that traits under selection are
linked to externalizing behaviour and risk-seeking. This might be partially
captured by our risk-aversion parameter $\sigma$, which affects fertility by
changing the marginal utility of income, but a more direct channel is risky
sexual behaviour [@mills2021identification]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and younger age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures such as waist-hip ratio and BMI. We
test this theory directly by re-estimating equation \@ref(eq:regression)
controlling for a measure of risk attitude (UK Biobank field 2040).  The median
ratio of effect sizes between regressions with and without controls is 
`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
uncontrolled regressions remain so when controlling for risk attitude. This
non-result could simply reflect the imprecision of the risk aversion measure,
which is a single yes/no question. However, the risk aversion measure *does*
predict the overall number of children, highly significantly ($p < 2 \times
10^{-16}$ in `r n_risk_sig` out of 33 regressions). Given that, and the statistical power
we get from our sample size, we believe that the non-result is real: while risk
aversion does predict fertility in the sample, it is not an important channel
for natural selection.

@@

\FloatBarrier
\clearpage
## Parents' generation

@@

Excluding childless people in the parents' generation could bias our estimates.
To learn about this, we compare effect sizes excluding and including childless
people in the *current* generation. The correlation
between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
different scores are broadly similar whether the childless are counted or not. 
However, absolute effect sizes are smaller when the
childless are excluded, for `r n_smaller` out of 33 scores; the median
percentage change is `r percent(median_pct_diff)`. The fact that childless
people have such a strong effect on estimates makes it hard to compare total
effect sizes across generations. In particular, since the parents' generation has
a different distribution of numbers of children, childless people may have had 
more or less effect in that generation.

@@

>>>>

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals of European descent, corrected for ancestry using 100 genetic
principal components (see Materials and Methods). Figure
\@ref(fig:plot-means-over-time) plots mean polygenic scores in the sample by
5-year birth intervals. Several scores show consistent increases or declines
over this 30-year period, of the order of 5% of a standard deviation. These
changes could reflect natural selection within the UK population, but also
emigration, or ascertainment bias within the sample. Respondents have higher
income and are better educated than the UK population, and they may also differ
on other unobserved characteristics [@10.1093/aje/kwx246]. Since richer and more
educated people also live longer, this bias could also increase with age.
Lastly, the sample sex ratio skews `r female_ratio * 100`% female.

@@

To correct for ascertainment bias, we weight participants using population
data. We try three alternative weighting schemes: (1) "Geographical" weighting,
by geography, age and presence/absence of a partner; (2) age and highest educational
qualification; and (3) for women only, age, highest qualification, and age at 
first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes, and also with uniform
weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
reaching significance under at least one scheme. Weighting makes
a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic variables.[^weighting]

[^weighting]: To make interpretation straightforward, we do not use weights in 
results below unless noted. However, all our results are robust to using weights.

We also check for balancing and diversifying selection by estimating
\@ref(eq:regression) with a quadratic term.[^balance-diverse-def] In particular,
we find diversifying selection for educational attainment polygenic scores: at
higher values of these scores, the negative effect on fertility is smaller
(Appendix Figure \@ref(fig:plot-purifying)).

[^balance-diverse-def]: Balancing selection selects for intermediate values, 
reducing the variance of the trait under selection. Diversifying selection
selects for extreme values and increases the variance.

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income, education and sex, and
family structure variables including age at first live birth, presence of a partner, and
lifetime number of sexual partners.
We emphasize that these categories are not necessarily exogenous to the polygenic
scores. For example, both in the data and in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related 
polygenic scores. Nevertheless, differences in equation
\@ref(eq:regression) across subgroups constrain the set of possible explanations
for natural selection. As we show below, the economic theory predicts 
these differences, while alternative explanations do not.

@@

To show this, we write down a simple model of fertility choices. $h$ is
an individual's level of human capital. For now, we simply identify this with
his or her wage $W$. Raising a child takes time $b$. People maximize utility
from the number of children $N$ and from income $Y\equiv(1-bN)W$. An individual's
payoff is

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
predictions, along with corresponding empirical results for the correlation
between polygenic scores and fertility.[^one-period-results] The key insight of
the model is that for middling levels of risk aversion, the substitution effect
dominates at low income levels, but is balanced by the income effect as income
increases.

[^one-period-results]: Predictions 1-3 also hold in the one-period model with
CRRA utility. Our empirical results are actually stronger than prediction 5, in
that correlations with fertility are *reversed* at higher AFLB. This prediction
can be accommodated in the model if children have a money cost as well as a time
cost. Appendix Figure \@ref(fig:N-plot-with-m) shows an example.

@@

The above does not aim to provide a complete theory of fertility. Rather, it
shows that a simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise.

@@

We consider three alternative theories that might explain our results. The first
is that welfare benefits which incentivize child-bearing might be taken up
among low-income individuals. However, our effect sizes appear unchanged
over a large span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and 
\@ref(tab:tbl-children-by-period)), during which
government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter $a$
in the model [cf. @jones2008fertility]. This theory would not explain why our
results are weaker at higher incomes and education levels.[^note] A third
alternative is that traits under selection are linked to externalizing behaviour
and risk-seeking. This might be partially captured by our risk-aversion
parameter $\sigma$, which affects fertility by changing the marginal utility of
income, but a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk aversion measure, which is a single yes/no question.
However, the risk aversion measure *does* predict the overall number of
children, highly significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of
33 regressions). Given that, and the statistical power we get from our sample
size, we believe that the non-result is real: while risk aversion does predict
fertility in the sample, it is not an important channel for natural selection.

@@

\FloatBarrier
\clearpage
## Parents' generation {#sec-parents}

@@

Excluding childless people in the parents' generation could bias our estimates.
To learn about this, we compare effect sizes excluding and including childless
people in the *current* generation. The correlation
between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
different scores are broadly similar whether the childless are counted or not. 
However, absolute effect sizes are smaller when the
childless are excluded, for `r n_smaller` out of 33 scores; the median
percentage change is `r percent(median_pct_diff)`.

@@

The fact that childless people have such a strong effect on estimates makes it
hard to compare total effect sizes across generations. In particular, since the
parents' generation has a different distribution of numbers of children,
childless people may have had more or less effect in that generation. Another
issue is that we are in effect estimating parents' polygenic scores by the
scores of their children. This introduces some noise into our independent
variable, which might lead to errors-in-variables and bias coefficients towards
zero.
